Covering Your Basis: Model Agnostic Probes of the Binary Black Hole Population

### 2022-09-25 src/tex/methods.tex
methods.tex is mostly finished but w/o citations

A common non-parametric method in statistical applications is the use of basis splines. A spline function of order $k$, 
is a piece-wise polynomial of order $k$ polynomials stitched together from defined ``knot'' locations across the domain. 
They provide a useful and cheap way to interpolate generically smooth functions from a finite sampling of ``knot'' heights. 
Basis splines of order $k$ are a set of order $k$ polynomials that form a complete basis for any spline function of order $k$. 
Given an array of knot locations, $\mathbf{t}$ or knot vector, there exists a single unique linear combination of B-Splines for 
every possible spline function with knots, $\mathbf{t}$. One can construct the B-Splines basis components for a given knot vector 
using the Cox-de Boor recursion formula. Given knots $t_0$, $t_1$,...,$t_{i+k}$ we start with the base case as:

@@

@@

\noindent allowing one to build a basis of $i$ components that spans the vector space of order-$k$ spline functions interpolated 
from this knot vector. This is known as the ``B-Spline'' basis.

A similar basis, the M-Spline basis, has desirable properties when looking to model a probability density function. Namely, 
$M_i \geq 0$ within $(t_i, t_i+k)$, zero elsewhere, and normalized to unity, $\int M_i(x)dx = 1$. The M-Spline basis only differs 
from B-Splines by a normalization factor, namely:

\noindent We use this definition of the M-Spline basis for the models presented and applied in the later sections of this paper.

@@

\bruce{placeholder paragraph here to introduce our model constructions.}
We use hierarchical Bayesian inference to simultaneosly infer the mass, spin and redshift distributions of binary black holes (BBHs) using 
the 69 confident BBH detections reported in the recent LVK catalog of observations, GWTC-3. In particular we choose to parameterize the two
masses with the primary (more massive component) mass and the mass ratio defined as $q=\frac{m_2}{m_1}$. We choose to model 4 of the 6 total 
spin degrees of freedom of a binary merger with each component spin magnitude, and the cosine of the tilt angle which is defined as the angle 
between each spin vector and the binaries orbital angular mommentum vector. We additionally will fit a population model on the redshift distribution
of BBHs. We assume a $\Lambda\mathrm{CDM}$ cosmology defined by the Planck 2015 results, which defines an analytical mapping between each event's inferred
luminosity distance and it's redshift. While we choose to use MSpline distibutions for the masses and spins, we use the canonical \textsc{PowrerlawRedshift}
model describing the evolution of CBC merger rates as a power law in $(1+z)$ with a slope of $\lambda_z$. The corresponsing distribution is Thus
desribed by $p(z|\lambda)\propto \frac{dV_c}{dz} \frac{T_\mathrm{obs}}{1+z} (1+z)^\lambda$. We model the spin magnitude and tilt distibutions to be 
described by independent MSpline distributions for each component.

>>>>

A common non-parametric method in statistical applications is the use of basis splines \bruce{cite}. A spline function of order $k$, 
is a piece-wise polynomial of order $k$ polynomials stitched together from defined ``knot'' locations across the domain. 
They provide a useful and cheap way to interpolate generically smooth functions from a finite sampling of ``knot'' heights. 
Basis splines of order $k$ are a set of order $k$ polynomials that form a complete basis for any spline function of order $k$. 
Given an array of knot locations, $\mathbf{t}$ or knot vector, there exists a single unique linear combination of B-Splines for 
every possible spline function with knots, $\mathbf{t}$. One can construct the B-Splines basis components for a given knot vector 
using the Cox-de Boor recursion formula \bruce{cite}. Given knots $t_0$, $t_1$,...,$t_{i+k}$ we start with the base case as:

@@

\noindent allowing one to build a basis of $i$ components that spans the vector space of order-$k$ spline functions interpolated 
from this knot vector. This is known as the ``B-Spline'' basis after it's inventor de Boor \bruce{cite}. The power of basis splines
comes from the fact that you only have to do a somewhat expensive interpolation once for each set of points you want to evaluate your
spline functions at. This allows for much quicker evaluation time rather than having to re-do a \bruce{Is this right???} $\mathcal{O}(n^2)$
operation one can do a simple dot product of 2d matrix and 1d vector which is easily  $\mathcal{O}(n)$. This means it is much more feasible 
to use large numbers of knots and GPU calculations for expensive markov chain monte carlo (MCMC) based analyses.

A similar basis, the M-Spline basis \bruce{cite}, has desirable properties when looking to model a probability density function, which is the 
goal of population inference. Namely, $M_i \geq 0$ within $(t_i, t_i+k)$, zero elsewhere, and normalized to unity, $\int M_i(x)dx = 1$. We now have
a useful basis that can represent all spline functions that are probability density functions given our array of knots. The M-Spline basis differs 
from B-Splines by the normalization factor:

@@

\noindent We use this definition of the M-Spline basis for the models presented and applied in the later sections of this paper. 
Figure \ref{fig:spline_basis} shows an example of a 3rd order M-Spline basis with 20 degrees of freedom and knots linearly spaced from 0 to 1.
Another important feature of basis splines is that under appropiate prior conditions, one can alleviate sensitivities to arbitrarily 
chosen prior specifications that splines commonly struggle with. We look to the use of penalized splines or P-Splines \bruce{cite P-splines}, 
that adds a smoothing penalty (or prior) based on the difference of basis spline coeficients. This allows one to much more denseley populate
the domain with knots without the worry of extra variance in the inferred spline functions. We discuss the details of this smoothing prior 
in Appendix \ref{sec:psplines}.

@@

\bruce{placeholder paragraph here to introduce our model constructions.}
We use hierarchical Bayesian inference to simultaneosly infer the astrophysical mass, spin and redshift distributions of 
a binary black holes (BBHs) given a catalog of gravitational wave observations. See Appendix \ref{sec:hierarchical_inference} for a detailed 
description of the analysis framework. We choose to parameterize the binaries' masses with the primary (defined as the more massive component) mass 
and the mass ratio -- defined as $q=\frac{m_2}{m_1}$, bounded in the range from 0 to 1. We choose to model 4 of the 6 total 
spin degrees of freedom of a binary merger with each component spin magnitude $a_1$ and $a_2$, along with the cosine of the tilt angle of each component, 
$\cos{\theta_1}$ and $\cos{\theta_2}$. The tilt angle is defined as the angle between each components' spin vector and the binaries orbital angular mommentum vector. 
We additionally will fit a population model on the redshift or luminosity distance distribution of BBHs. We assume a $\Lambda\mathrm{CDM}$ cosmology using the 
parameters from the Planck 2015 results \bruce{cite here}, which defines an analytical mapping between each event's inferred
luminosity distance and it's redshift. While we choose to use MSpline distibutions for the primary mass and spin distributions, we 
model the mass ratio distribution with a power law , i.e. $p(q | m_1, m_\mathrm{min}, \beta_q) \propto q^{\beta_q} \Theta(qm_1 - m_\mathrm{min}) \Theta(m_1 - qm_1)$. 
Where $\Theta$ is the Heaviside step function that ensures $m_2$ is within the range [$m_\mathrm{min}$, $m_1$].
We model the evolution of the merger rate with redshift with the \textsc{PowrerlawRedshift} \bruce{cite mayas paper} model, which is desribed by 
$p(z|\lambda_z)\propto \frac{dV_c}{dz} \frac{T_\mathrm{obs}}{1+z} (1+z)^\lambda_z$. We describe our specific prior choices on the hyper-parameters and each model 
in more detail in Appendix \ref{sec:modelpriors}.
